\hspace{1cm} Do syntactic and semantic similarity lead to interference effects?\newline Evidence from self-paced reading and event-related potentials using German

### 2023-10-15 paper/main.tex
finished one pass through paper

\authornote{This research was funded by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation), Project ID 317633480, SFB 1287.
We thank Johanna Thieke, Romy Leue, Elise Oltrogge and Lisa Plagemann for assistance during data acquisition.
Correspondence concerning this article should be addressed to Pia Schoknecht, e-mail: pia.schoknecht@uni-potsdam.de
}

\abstract{Cue-based retrieval accounts of sentence processing postulate that at a verb, retrieval cues are generated to complete a dependency with the verb's argument(s); for example, the dependency between the subject and the verb must be completed to interpret the sentence. If these retrieval cues match with not only the features on the noun that is the retrieval target but also with those on other nouns in the sentence, then processing difficulty arises at the verb. This difficulty in identifying the correct dependent is called similarity-based interference. 
% We are explaining what we did, not summarizing every detail in the paper:
%Previously, behavioral studies have consistently found syntactic interference at the verb, but semantic interference was only found in later regions or not at all. 
% What we did:
We present large-sample self-paced reading (N = 774) and event-related potentials experiments (N = 103) using a well-established design \citep{vandyke07}, investigating interference due to syntactic and semantic cues in German. 
% Not needed for an abstract, plus sounds boring:
%The aim of our experiments was to gain precise estimates of the effects of syntactic and semantic interference to inform theory building.
% What we found:
A Bayes factor analysis showed that 
both the self-paced reading times as well as event-related potentials provided clear evidence for a semantic interference effect. In the self-paced reading experiment, interference due to the semantic cues caused a reading time slow-down starting at the distractor. 
% The reader does not even know what that is:
%The best explanation for this effect is encoding interference. 
The event-related potentials showed a more negative N400 amplitude at the critical verb, i.e., the retrieval site, for high vs.\ low semantic interference.
% Irrelevant in an abstract:
 %and no such effects at preceding words. 
 % No new information in this sentence, a waste:
 %Therefore, the event-related potentials were modulated by retrieval interference.
% Said this more compactly already:
%Bayes factors for both reading times and event-related potentials provided evidence for the semantic interference effect. 
Surprisingly, in both experiments, Bayes factor analyses showed evidence against interference due to syntactic cues. This finding contradicts the predictions of the standard implementations of cue-based retrieval theory, which (implicitly) assume that both syntactic and semantic cues play a role in retrieval. We show through computational modeling that cue-based retrieval also shows no syntactic interference if the parser is assumed to keep track of which clause a noun occurs in. Thus, if syntactic retrieval cues include hierarchical syntactic information (is the noun in the same clause as the verb?), the predictions of the cue-based retrieval model are consistent with the observed patterns in our data.

In sum, this large-sample study of German using self-paced reading and event-related potentials suggests that although both syntactic and semantic cues drive retrieval at the verb, syntactic cues include hierarchical syntactic information, which can render the parser immune to syntactic interference effects.

@@

Thus, one theoretical assumption that would be consistent with the observed SPR and EEG data is that the parser does not treat every subject in the sentence equally as a possible retrieval candidate; only the relevant subjects (those in the same clause) are considered.  This is of course a speculative claim that would need to be further investigated empirically.

@@

The comparison of the results of our SPR and ERP experiments suggests that the SPR results were strongly affected by encoding interference, while the ERPs elicited by the critical verb were not. This difference between methods might be caused by the difference in presentation rate. In the self-paced reading experiment, as implied by the name, the participants had control over the presentation rate of the stimuli. The observed slowdown starting at the distractor  might have resulted from the probably subconscious intent to slow down the sampling rate of the incoming stimuli to gain more time to process them. In contrast, in the ERP experiment, we used rapid serial visual presentation (RSVP), which presents stimuli at a fixed rate that is beyond the control of the participant. Therefore, they must adapt to the set rate and process the stimuli as fast as the rate requires. Here, it is noteworthy that the presentation duration of 500\,ms per word and 400\,ms between words which was used in the present ERP experiment is slower than usual natural reading speed. However, while in SPR the slowdown due to encoding and maintaining the distractor in memory overshadowed potential retrieval effects, the procedure of the ERP experiment prevented that. In the ERP results, we observed an effect at the critical verb, which can be attributed to memory retrieval of the subject. This retrieval was only affected by the experimental manipulation.

\subsection{Relation to other sentence processing accounts}
Besides our considerations above regarding the possibility that another syntactic cue (e.g., \{± same clause\}) was used, our results highlight the importance of the semantic cue for subject-verb dependency resolution.

The evidence against a main effect of syntactic interference and hence the relative unimportance of the syntactic cue \{± grammatical subject\} for subject-verb dependency resolution in our study stands in contrast to predictions from syntax-first accounts \citep[e.g.,][]{frazier_clifton1996, frazier1979_diss, frazier1987}. These accounts predict that syntactic constraints are considered first during sentence processing. At least, our results speak against syntax as a strong filter. However, we found that a semantically similar distractor induced interference. This finding can be easily reconciled with various accounts of sentence processing. Above we have already discussed how our results relate to predictions from the \citet{Lewis2005} cue-based retrieval model. Our results fit predictions from constraint-based models \citep[e.g.,][]{macdonald1994, mcrae1998modeling, trueswell1993}. Under constraint-based accounts, it is assumed that multiple constraints simultaneously and immediately influence processing. These accounts predict our ERP results: Compared to an inanimate distractor, an animate distractor -- regardless of its syntactic status -- would impede processing at a verb which requires an animate subject. The lack of a syntactic interference effect could indicate that the status `grammatical subject' was not used as a syntactic constraint. As discussed above, a plausible syntactic constraint is that a verb and its arguments need to be within the same clause. Furthermore, our results are in line with the good-enough processing account \citep{ferreira2007goodenough}. This account assumes that comprehenders do not always aim to build a fully fleshed-out analysis of a sentence, but instead might accept incomplete or even incorrect analyses. The high number of participants that needed to be excluded in our experiments due to accuracy below 70\,\% (117 out of 908 participants in the SPR experiment, 29 out of 146 participants in the EEG experiment), suggests that our materials posed high processing demands. It is reasonable to assume that the participants might have sometimes adopted a good-enough processing mode to manage resources. So, given a good-enough processing mode which leaves some syntactic relations within the sentence underspecified, it makes sense for the comprehender to use a simple heuristic for subject-verb dependency resolution. Animacy is an obvious choice for such a heuristic. Animate entities are proto-typical agents \citep{dowty1991thematic} and therefore, the primary use of the \{+ animate\} cue to retrieve a subject probably leads to the correct analysis (not just within the experiment context, but in everyday language use). Finally, our findings support the possibility that language processing relies predominantly on semantic associations to form (probabilistic)
representation of event structures. This view has been put forward by \citet{rabovsky_etal_2018}, when they used the sentence gestalt model \citep{mcclelland1989_sentence_gestalt}, a neural network model of sentence comprehension, to model the N400 amplitude. This model assumes that language comprehension relies on associative form-to-meaning mapping instead of syntactic rules. This is in line with our results which highlight the importance of semantic features such as animacy over syntactic functions such as grammatical subject for sentence processing.

@@

This project investigated the use of  syntactic and semantic cues during subject-verb dependency resolution with self-paced reading and event-related potentials. While our results did not directly replicate previous results from eye-tracking while reading studies \citep{vandyke07, mertzen}, our SPR and ERP results were consistent. We attributed differences between the present and previous studies to differences in statistical power and experimental methodology. Both our experiments showed semantic interference, i.e., more processing difficulty during long-distance subject-verb dependency resolution in the presence of a semantically matching distractor. This manifested in longer self-paced reading times starting at the distractor and in a more negative N400 amplitude for the high vs.\ low semantic interference conditions elicited by the critical verb. We discussed that these different time courses in the two experiments are due to different types of interference: Encoding interference in the SPR experiment and retrieval interference in the ERP experiment. We did not find effects of syntactic interference. We assume that the syntactic cue \{± grammatical subject\} which was manipulated in our materials, was not used during sentence processing. We proposed that another syntactic cue \{± same clause\} might be used \citep[see also][]{mertzen}. We suggest that the present findings, which are in line with various sentence processing accounts, highlight the importance of semantic cues during sentence processing.

>>>>

\authornote{This research was funded by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation), Project ID 317633480, SFB 1287.
We thank Johanna Thieke, Romy Leue, Elise Oltrogge and Lisa Plagemann for assistance during data acquisition. We are grateful to Dario Paape for helpful discussions, and Himanshu Yadav for assistance with the computational modeling. 
Correspondence concerning this article should be addressed to Pia Schoknecht, e-mail: pia.schoknecht@uni-potsdam.de
}

\abstract{Cue-based retrieval accounts of sentence processing postulate that at a verb, retrieval cues are generated to complete a dependency with the verb's argument(s); for example, the dependency between the subject and the verb must be completed to interpret the sentence. If these retrieval cues match with not only the features on the noun that is the retrieval target but also with those on other nouns in the sentence, then processing difficulty arises at the verb. This difficulty in identifying the correct dependent is called similarity-based interference. 
% We are explaining what we did, not summarizing every detail in the paper:
%Previously, behavioral studies have consistently found syntactic interference at the verb, but semantic interference was only found in later regions or not at all. 
% What we did:
We present large-sample self-paced reading (N = 774) and event-related potentials experiments (N = 103) using a well-established design \citep{vandyke07,mertzen}, to investigate interference due to syntactic and semantic cues in German. 
% Not needed for an abstract, plus sounds boring:
%The aim of our experiments was to gain precise estimates of the effects of syntactic and semantic interference to inform theory building.
% What we found:
A Bayes factor analysis showed that 
both the self-paced reading times as well as event-related potentials provided clear evidence for a semantic interference effect. In the self-paced reading experiment, interference due to the semantic cues caused a reading time slow-down starting at the distractor. 
% The reader does not even know what that is:
%The best explanation for this effect is encoding interference. 
The event-related potentials showed a more negative N400 amplitude at the critical verb, i.e., the retrieval site, for high vs.\ low semantic interference.
% Irrelevant in an abstract:
 %and no such effects at preceding words. 
 % No new information in this sentence, a waste:
 %Therefore, the event-related potentials were modulated by retrieval interference.
% Said this more compactly already:
%Bayes factors for both reading times and event-related potentials provided evidence for the semantic interference effect. 
Surprisingly, in both experiments, Bayes factor analyses showed evidence against interference due to syntactic cues. This finding contradicts the predictions of the standard implementations of cue-based retrieval theory, which (implicitly) assume that both syntactic and semantic cues play an equal role in retrieval. We show through computational modeling that cue-based retrieval also shows no syntactic interference if the parser is assumed to keep track of which clause the subject occurs in. Thus, if syntactic retrieval cues include hierarchical syntactic information (is the noun in the same clause as the verb?), the predictions of the cue-based retrieval model would be consistent with the observed patterns in our data.

In sum, this large-sample study of German using self-paced reading and event-related potentials suggests that although both syntactic and semantic cues drive retrieval at the verb, syntactic cues may include hierarchical syntactic information, which can render the parser immune to syntactic interference effects.

@@

Thus, one theoretical assumption that would be consistent with the observed SPR and EEG data is that the parser uses the animacy cue exactly as proposed in previous work on cue-based retrieval; however, the parser uses the syntactic cue while keeping track of the syntactic location of the subject.  This is of course a speculative claim that would need to be further investigated empirically.

@@

The comparison of the results of our SPR and ERP experiments suggests that the SPR results were strongly affected by encoding interference, while the ERPs elicited by the critical verb were not. This difference between methods might be caused by the difference in presentation rate. In the self-paced reading experiment, as implied by the name, the participants had control over the presentation rate of the stimuli. The observed slowdown starting at the distractor  might have resulted from the probably subconscious intent to slow down the sampling rate of the incoming stimuli to gain more time to process them. In contrast, in the ERP experiment, we used rapid serial visual presentation (RSVP), which presents stimuli at a fixed rate that is beyond the control of the participant. Therefore, they must adapt to the set rate and process the stimuli as fast as the rate requires. Here, it is noteworthy that the presentation duration of 500\,ms per word and 400\,ms between words which was used in the present ERP experiment is slower than usual natural reading speed. However, while in SPR the slowdown due to encoding and maintaining the distractor in memory overshadowed potential retrieval effects, the procedure of the ERP experiment prevented that. In the ERP results, we observed an effect at the critical verb, which can be attributed to the retrieval process hypothesized by cue-based retrieval accounts.

\subsection{Relation to other sentence processing accounts}

%Besides our considerations above regarding the possibility that another syntactic cue (e.g., \{± same clause\}) was used, our results highlight the importance of the semantic cue for subject-verb dependency resolution.

The evidence against a main effect of syntactic interference for subject-verb dependency resolution in our study is not consistent with the assumption in previous work, e.g., \cite{vandyke07,mertzen}, that the parser searches for a subject simply by setting the retrieval cue \{± grammatical subject\}.
  However, the semantic interference effect we found can be easily reconciled with previous work in  sentence processing. In the modeling section above, we have already discussed how our results relate to predictions from the \citet{Lewis2005} cue-based retrieval model.  
  %Our results may also be consisted with constraint-based models \citep[e.g.,][]{macdonald1994, mcrae1998modeling, trueswell1993}. Under constraint-based accounts, it is assumed that multiple constraints simultaneously and immediately influence processing. These accounts predict our ERP results: Compared to an inanimate distractor, an animate distractor -- regardless of its syntactic status -- would impede processing at a verb which requires an animate subject. The lack of a syntactic interference effect could indicate that the status `grammatical subject' was not used as a syntactic constraint. As discussed above, a plausible syntactic constraint is that a verb and its arguments need to be within the same clause. 
 Our results could also be seen as consistent with the good-enough processing account \citep{ferreira2007goodenough}, which assumes that comprehenders do not always aim to build a fully fleshed-out analysis of a sentence, but instead might accept incomplete, underspecified, or even incorrect analyses. The high number of participants that needed to be excluded in our experiments due to accuracy below 70\,\% (117 out of 908 participants in the SPR experiment, 29 out of 146 participants in the EEG experiment), suggests that our materials led to high processing demands. It is reasonable to assume that the participants might have--at least occasionally--adopted a good-enough processing mode when faced with high task demands \parencite{swets2008underspecification,LogacevMultiple} or due to working memory capacity limitations \parencite{MalsburgVasishth2013}, or both. So, given a good-enough processing mode which leaves some syntactic relations within the sentence underspecified, it makes sense for the comprehender to use a simple heuristic for subject-verb dependency resolution. Animacy is an obvious choice for such a heuristic. Animate entities are proto-typical agents \citep{dowty1991thematic} and therefore, the primary use of the \{+ animate\} cue to retrieve a subject leads to the correct analysis with high probability (not just within the experiment context, but in everyday language use). Finally, our findings are also consistent with the possibility that language processing relies predominantly on semantic associations to form (probabilistic)
representation of event structures. This view has been put forward by \cite{rabovsky_etal_2018}, when they used the neural-network sentence gestalt model \parencite{mcclelland1989_sentence_gestalt}, to model the N400 amplitude. This model assumes that language comprehension relies on associative form-to-meaning mapping instead of syntactic rules. This assumption is consistent with our results.

Without the modeling results reported above, it would be easy to conclude that our data show that  syntactic cues play no role in sentence processing. However, this conclusion would be obviously wrong. First, as we showed above, the retrieval model suggests that--at least in our data--the subject cue is used, but differently than previously assumed: the relevant cue is subject-in-same-clause. Second, there is plenty of independent evidence for the central role of syntactic information in comprehension; some examples are the role of case marking \parencite{ALV2019,HusainEtAl2014}, syntactic constraints like Principle A of the binding theory \parencite{Sturt2003,dillon2013,yadav2021individual},  and 
the importance of word order \parencite{meng2000mode}. There are of course important examples of the parser ignoring syntactic information. Three dramatic examples are the grammaticality illusion in multiple embeddings \parencite{gibsonthomas99}, the agreement attraction phenomenon \parencite{wagersetal,lago_etal_2021,jaeger_etal_2017}, and local coherence effects \parencite{taboretal04}.  However, there are languages (such as German and Dutch) in which the grammaticality illusion does not occur \parencite{VSLK11,FrankEtAl2015}, and a large-sample study \parencite{lcpaape2023} presents evidence against the syntactic local coherence effect that was reported by \cite{taboretal04}. Thus, it seems that the role of syntax in building incremental structure remains as important as ever.

@@

This project investigated the use of  syntactic and semantic cues during subject-verb dependency resolution with self-paced reading and event-related potentials.  Both our experiments consistently showed semantic interference, i.e., more processing difficulty during long-distance subject-verb dependency resolution in the presence of a semantically matching distractor. This manifested in longer self-paced reading times starting at the distractor and continuing into the post-critical region, and in a more negative N400 amplitude for the high vs.\ low semantic interference conditions elicited by the critical verb. These different time courses in the two experiments are likely to be due to different types of interference: Encoding interference in the SPR experiment and retrieval interference in the ERP experiment. These differences are likely to be due to the methods used (self-paced vs.\ auto-paced reading).Overall, we found evidence against syntactic interference; computational modeling shows that the most likely explanation for this is that the parser uses the syntactic cue  \{± subject-in-same-clause\}  to identify the correct target for retrieval. A broader implication is that the parser may be able to use hierarchical structure in the input to intelligently target the correct syntactic dependent from memory.
